---
title: "Plots"
author: "Kapil Parab/Nicaise Irambona"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing libraries

```{r}
library(sf)
library(tigris)
library(tidyverse)
library(ggrepel)
library(ggforce)
library(scales)
```

```{r}
# US States
us_states <- st_read("data/us_states/tl_2024_us_state.shp") %>%
  st_transform(., crs = 4326)

# Getting Arizona state details
az_counties <- counties("AZ", year = 2024) %>% st_transform(., crs = 4326)

az_landmarks <- landmarks("AZ", year = 2024) %>% st_transform(., crs = 4326)

az_road <- primary_secondary_roads("AZ", year = 2024) %>% st_transform(., crs = 4326)

# City outlines for Arizona
az_city_outline <- st_read("data/arizona_city_boundary/arizona_city.shp") %>%
  st_transform(., crs = 4326)
```

# Reading dataset

```{r}
vehicles <-
  list.files(path = "./data/cleansed/vehicles/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ read_csv(., show_col_types = FALSE))

motorcycles <-
  list.files(path = "./data/cleansed/motorcycles/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ read_csv(., show_col_types = FALSE))
```

# Visualizing vehicle accidents across US

```{r}
state_accidents <- vehicles %>%
  select(CASE_ID, STATE) %>%
  distinct() %>%
  group_by(STATE) %>%
  summarise(COUNT = n())

us_states_veh <- inner_join(us_states, state_accidents, by = c("NAME" = "STATE"))

us_state_veh_label <- tail(us_states_veh[order(us_states_veh$COUNT), ], 15)

ggplot(us_states_veh) +
  geom_sf(aes(fill = COUNT), color = "white") +
  coord_sf(xlim = c(-70, -170)) +
  geom_label_repel(
    data = us_state_veh_label,
    aes(label = NAME, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  theme_bw() +
  labs(title = "Vehicle Accidents across United States", subtitle = "2020-2022") +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "inside",
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.position.inside = c(0.8, 0.82),
    legend.spacing.x = unit(0.5, "cm"),
    legend.background = element_blank(),
    plot.background = element_rect(fill = "white")
  ) +
  guides(fill = guide_colorbar(
    title = "Count of Accidents",
    label.position = "bottom",
    title.position = "top",
    title.vjust = 1,
    frame.colour = "black",
    barwidth = 8,
    barheight = 1.5
  ))

ggsave("veh_acc_us.png", path = "plots")
```
# Visualizing motorcycle accidents across US

```{r}
state_accidents <- motorcycles %>%
  select(CASE_ID, STATE) %>%
  distinct() %>%
  group_by(STATE) %>%
  summarise(COUNT = n())

us_states_moto <- inner_join(us_states, state_accidents, by = c("NAME" = "STATE"))

us_state_moto_label <- tail(us_states_moto[order(us_states_moto$COUNT), ], 15)

ggplot(us_states_moto) +
  geom_sf(aes(fill = COUNT), color = "white") +
  coord_sf(xlim = c(-70, -170)) +
  geom_label_repel(
    data = us_state_moto_label,
    aes(label = NAME, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  theme_bw() +
  labs(title = "Motorcycle Accidents across United States", subtitle = "2020-2022") +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "inside",
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.position.inside = c(0.8, 0.82),
    legend.spacing.x = unit(0.5, "cm"),
    legend.background = element_blank(),
    plot.background = element_rect(fill = "white")
  ) +
  guides(fill = guide_colorbar(
    title = "Count of Accidents",
    label.position = "bottom",
    title.position = "top",
    title.vjust = 1,
    frame.colour = "black",
    barwidth = 8,
    barheight = 1.5
  ))

ggsave("veh_moto_us.png", path = "plots")
```

# Filtering dataset to visualize Arizona accidents

```{r}
crash_moto_az <- motorcycles %>%
  filter(., STATE == "Arizona") %>%
  select(CASE_ID, ACC_TS, VEH_BRAND, LATITUDE, LONGITUDE) %>%
  distinct()

crash_veh_az <- vehicles %>%
  filter(., STATE == "Arizona") %>%
  select(CASE_ID, ACC_TS, VEH_CATEGORY, LATITUDE, LONGITUDE) %>%
  distinct()

az_city_labels <- tail(
  az_city_outline[order(az_city_outline$Shape__Are), ],
  10
)
```

# Plotting roadways and city outlines for Arizona

```{r}
ggplot() +
  geom_sf(data = az_counties, fill = "white") +
  geom_sf(data = az_city_outline, fill = "green") +
  geom_sf(data = az_road, color = "black") +
  geom_label_repel(
    data = az_city_outline,
    aes(label = Name, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  labs(title = "Cities and Major Roadways of Arizona") +
  theme(axis.title = element_blank(), axis.ticks = element_blank())

ggsave("az_city_roadways.png", path = "plots")
```

# Plotting vehicle crashes for Arizona

```{r}
ggplot() +
  geom_sf(data = az_counties, fill = "white") +
  geom_sf(data = az_city_outline, fill = "green") +
  geom_sf(data = az_road, color = "black") +
  geom_point(data = crash_veh_az, aes(x = LONGITUDE, y = LATITUDE), size = 1, color = alpha("red", 0.2)) +
  geom_label_repel(
    data = az_city_labels,
    aes(label = Name, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  labs(title = "Vehicle Crashes in Arizona", subtitle = "2020-2022") +
  theme(axis.title = element_blank(), axis.ticks = element_blank())

ggsave("az_veh_acc.png", path = "plots")
```

# Plotting motorcycle crashes for Arizona

```{r}
ggplot() +
  geom_sf(data = az_counties, fill = "white") +
  geom_sf(data = az_city_outline, fill = "green") +
  geom_sf(data = az_road, color = "black") +
  geom_point(data = crash_moto_az, aes(x = LONGITUDE, y = LATITUDE), size = 1, color = alpha("red", 0.2)) +
  geom_label_repel(
    data = az_city_labels,
    aes(label = Name, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  labs(title = "Motorcycle Crashes in Arizona", subtitle = "2020-2022") +
  theme(axis.title = element_blank(), axis.ticks = element_blank())


ggsave("az_moto_acc.png", path = "plots")
```

# Zooming in into vehicle crashes of Tucson

```{r}
tucson_outline <- az_city_outline %>% filter(., Name == "Tucson")

tucson_crash <- merge(crash_veh_az, crash_moto_az)

ua_outline <- az_landmarks %>% filter(., FULLNAME == "Univ of Arizona" & POINTID == 1101255267590)
```

```{r}
ggplot() +
  geom_sf(data = tucson_outline, fill = "grey") +
  geom_sf(data = ua_outline, fill = "blue") +
  geom_sf(data = az_road, color = "black") +
  geom_point(data = crash_veh_az, aes(x = LONGITUDE, y = LATITUDE), size = 1, color = alpha("red", 0.5)) +
  geom_point(data = crash_moto_az, aes(x = LONGITUDE, y = LATITUDE), size = 1, color = alpha("green", 0.5)) +
  geom_label_repel(
    data = ua_outline,
    aes(label = FULLNAME, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    max.iter = 10000,
    max.time = 10,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0.1,
    segment.color = "gray",
    force = 2,
    max.overlaps = Inf
  ) +
  labs(title = "Vehicle & Motorcycle Crashes in Arizona", subtitle = "2020-2022") +
  theme_bw() +
  theme(axis.title = element_blank(), axis.ticks = element_blank()) +
  coord_sf(xlim = c(-111.2, -110.7), ylim = c(31.98, 32.4))

ggsave("tucson_acc.png", path = "plots")
```

# Visualizing accident trend based on Time of Day

```{r}
crash_veh_hr <- crash_veh_az %>%
  mutate(ACC_HR = hour(ACC_TS)) %>%
  select(ACC_HR) %>%
  mutate(CAT = "Vehicles") %>%
  group_by(ACC_HR, CAT) %>%
  summarise(total_count = n())

crash_moto_hr <- crash_moto_az %>%
  mutate(ACC_HR = hour(ACC_TS)) %>%
  select(ACC_HR) %>%
  mutate(CAT = "Motorcycles") %>%
  group_by(ACC_HR, CAT) %>%
  summarise(total_count = n())

hr_merge <- merge(crash_veh_hr, crash_moto_hr, all = TRUE)

ggplot(data = hr_merge, aes(x = ACC_HR, y = total_count, group = CAT, color = CAT)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:23, name = "Hour") +
  scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100, 150, 200, 250, 300, 350, 400, 450, 500), name = "Count") +
  labs(y = "Total Count", color = "Category") +
  theme_bw() +
  theme(
    legend.position = "inside",
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.position.inside = c(0.19, 0.9),
    legend.background = element_rect(fill = "lightgrey"),
    axis.line = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Accidents in Arizona based on Time of Day", subtitle = "2020-2022"
  )

ggsave("az_tod.png", path = "plots")
```

# Visualizing accident trend based on Day of Month

```{r}
crash_veh_dt <- crash_veh_az %>%
  mutate(ACC_DATE = day(ACC_TS)) %>%
  select(ACC_DATE) %>%
  mutate(CAT = "Vehicles") %>%
  group_by(ACC_DATE, CAT) %>%
  summarise(total_count = n())

crash_moto_dt <- crash_moto_az %>%
  mutate(ACC_DATE = day(ACC_TS)) %>%
  select(ACC_DATE) %>%
  mutate(CAT = "Motorcycles") %>%
  group_by(ACC_DATE, CAT) %>%
  summarise(total_count = n())

dt_merge <- merge(crash_veh_dt, crash_moto_dt, all = TRUE)

ggplot(data = dt_merge, aes(x = ACC_DATE, y = total_count, group = CAT, color = CAT)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:31, name = "Day") +
  scale_y_continuous(breaks = c(0, 50, 100, 150, 200, 250), name = "Count") +
  labs(y = "Total Count", color = "Category") +
  theme_bw() +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    legend.title.position = "top",
    legend.position.inside = c(0.19, 0.9),
    legend.background = element_rect(fill = "lightgrey"),
    axis.line = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  labs(
    title = "Accidents in Arizona based on Day of Month", subtitle = "2020-2022"
  )

ggsave("az_dom.png", path = "plots")
```

# Probability of Survival Given Airbag Deployment and Location

```{r}
harm_event_fil <- c("Pedestrian", "Pedalcyclist", "Fell/Jumped from Vehicle", "Unknown")
col_filter <- c("Cargo/Vehicle Parts Set-In-Motion", "Non-Collision", "Not Reported", "Other Objects or Person Set-In-Motion", "Reported as Unknown", "Undercarriage")

airbag_data <- vehicles %>%
    filter(!is.na(AIRBAG_DEPLOYED) & 
           !is.na(AIRBAG_LOC) & 
           !is.na(INJURY) &
           !HARM_EV_TYPE %in% harm_event_fil &
           !IMPACT_LOC %in% col_filter
   ) %>%
  select(PERSON_TYPE, IMPACT_LOC, AIRBAG_DEPLOYED, AIRBAG_LOC, INJURY) %>%
  mutate(AIRBAG_LOC = case_when(
    AIRBAG_LOC == "Curtain (roof)" ~ "Curtain",
    AIRBAG_LOC == "Other (Knee, air belt, etc.)" ~ "Other",
    AIRBAG_LOC == "Side (door, seatback)" ~ "Door, Seatback",
    .default = AIRBAG_LOC
  ),
  SURVIVAL = ifelse(INJURY != "Death", 1, 0)
)

driver_airbag_prob <- airbag_data %>%
  filter(AIRBAG_DEPLOYED == "Yes" & PERSON_TYPE == "Driver") %>%
  group_by(IMPACT_LOC, AIRBAG_LOC) %>%
  summarise(
    total_cases = n(),
    survival_cases = sum(SURVIVAL),
    prob_survival_given_airbag = mean(SURVIVAL)
  )

passenger_airbag_prob <- airbag_data %>%
  filter(AIRBAG_DEPLOYED == "Yes" & PERSON_TYPE != "Driver") %>%
  group_by(IMPACT_LOC, AIRBAG_LOC) %>%
  summarise(
    total_cases = n(),
    survival_cases = sum(SURVIVAL),
    prob_survival_given_airbag = mean(SURVIVAL)
  )

ggplot(driver_airbag_prob, aes(x = AIRBAG_LOC, y = prob_survival_given_airbag)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Probability of Survival Given Airbag Deployment and Location",
    subtitle = "Driver",
    x = "Airbag Location",
    y = "Probability of Survival"
  ) +
  facet_grid(. ~ IMPACT_LOC, scales = "free", space = "free") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    strip.text = element_text(size = 6),
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10, unit = "pt")
  )

ggsave("airbag_surv_driver.png", path = "plots", dpi = 350)

ggplot(passenger_airbag_prob, aes(x = AIRBAG_LOC, y = prob_survival_given_airbag)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Probability of Survival Given Airbag Deployment and Location",
    subtitle = "Passenger",
    x = "Airbag Location",
    y = "Probability of Survival"
  ) +
  facet_grid(. ~ IMPACT_LOC, scales = "free", space = "free") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    strip.text = element_text(size = 6),
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10, unit = "pt")
  )

ggsave("airbag_surv_pass.png", path = "plots", dpi = 350)
```
# Probability of Survival Given Presense of Helment and Impact Location

```{r}
harm_event_fil <- c("Unknown")
col_filter <- c("Cargo/Vehicle Parts Set-In-Motion", "Non-Collision", "Not Reported", "Other Objects or Person Set-In-Motion", "Reported as Unknown", "Undercarriage")

helmet_data <- motorcycles %>%
    filter(!is.na(HELMET_PRESENT) &
           !is.na(INJURY) &
           !HARM_EV_TYPE %in% harm_event_fil &
           !IMPACT_LOC %in% col_filter &
             HELMET_MISUSE == "No"
   ) %>%
  select(IMPACT_LOC, HELMET_PRESENT, INJURY) %>%
  mutate(SURVIVAL = ifelse(INJURY != "Death", 1, 0)
)

conditional_prob <- helmet_data %>%
  filter(HELMET_PRESENT == "Yes") %>%
  group_by(IMPACT_LOC) %>%
  summarise(
    prob_survival = mean(SURVIVAL)
  )

ggplot(conditional_prob, aes(x = IMPACT_LOC, y = prob_survival)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Probability of Survival Given Helment and Impact Location",
    x = "Impact Location",
    y = "Probability of Survival"
  ) +
  theme(
    axis.text.x = element_text(angle = 20)
  )

ggsave("helmet_surv.png", path = "plots")
```

# Distribution of crashes across all states based on weather

```{r}
View(vehicles %>% select(WEATHER) %>% distinct())
```

```{r}
weather_fil <- c("Unknown", "Other")

weather_by_state <- vehicles %>%
  filter(., ! WEATHER %in% weather_fil) %>%
  select(STATE, WEATHER) %>%
    group_by(STATE, WEATHER) %>%
    summarise(count = n())

ggplot(weather_by_state, aes(x = WEATHER, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Weather Conditions by State",
    x = "Weather Condition",
    y = "Count"
  ) +
  facet_wrap(vars(STATE), ncol = 10, scales = "free_y") +
  theme(
    axis.text.x = element_text(angle = 90),  # Rotate x-axis labels for clarity
    strip.text = element_text(size = 8)  # Adjust facet label size
  )

for(i in 1:6){
  
  ggplot(weather_by_state, aes(x = WEATHER, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Weather Conditions by State",
    x = "Weather Condition",
    y = "Count"
  ) +
  facet_wrap_paginate(vars(STATE), ncol = 3, nrow = 3, page = i, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for clarity
    strip.text = element_text(size = 8)  # Adjust facet label size
  )
  
  ggsave(paste0("weather_state_", i, ".png"), dpi = 350, path = "plots")

}

```

# Accidents Across Age Groups

## Vehicles

```{r}
# Define age groups
veh_data <- vehicles %>%
  mutate(Age_Group = case_when(
    AGE < 18 ~ "Under 18",
    AGE >= 18 & AGE <= 25 ~ "18-25",
    AGE >= 26 & AGE <= 35 ~ "26-35",
    AGE >= 36 & AGE <= 50 ~ "36-50",
    AGE > 50 ~ "Above 50",
    TRUE ~ "Unknown"
  )) %>%
  filter(Age_Group != "Unknown" & PERSON_TYPE == "Driver") %>%
  mutate(Age_Group = factor(Age_Group, levels = c("Under 18", "18-25", "26-35", "36-50", "Above 50")))
```

### Drinking

```{r}
veh_drink_total <- veh_data %>%
  filter(., DRINKING_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

veh_drink_death <- veh_data %>%
  filter(INJURY == "Death" & DRINKING_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

# Create a bar chart
ggplot() +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Total"), 
    data = veh_drink_total,
    stat = "identity", show.legend = FALSE
  ) +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Death"), 
    data = veh_drink_death,
    stat = "identity", show.legend = FALSE
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 10)  # Generate 10 evenly spaced breaks
  ) +
  scale_fill_manual(
    values = c("Total" = "#1fbcff", "Death" = "#ff661f"),  # Assign colors to categories
    labels = c("Total Count", "Deaths")              # Legend labels
  ) +
  labs(
    title = "Accidents Across Age Groups",
    subtitle = "Drinking and Death",
    x = "Age Group",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
  )

ggsave("veh_drink_death.png", path = "plots", dpi = 350)
```
### Drugs

```{r}
veh_drugs_total <- veh_data %>%
  filter(., DRUGS_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

veh_drugs_death <- veh_data %>%
  filter(INJURY == "Death" & DRUGS_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

# Create a bar chart
ggplot() +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Total"), 
    data = veh_drugs_total,
    stat = "identity", show.legend = FALSE
  ) +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Death"), 
    data = veh_drugs_death,
    stat = "identity", show.legend = FALSE
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 10)  # Generate 10 evenly spaced breaks
  ) +
  scale_fill_manual(
    values = c("Total" = "#1fbcff", "Death" = "#ff661f"),  # Assign colors to categories
    labels = c("Total Count", "Deaths")              # Legend labels
  ) +
  labs(
    title = "Accidents Across Age Groups",
    subtitle = "Drugs and Death",
    x = "Age Group",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
  )

ggsave("veh_drugs_death.png", path = "plots", dpi = 350)
```

## Motorcycles

```{r}
# Define age groups
moto_data <- motorcycles %>%
  mutate(Age_Group = case_when(
    AGE < 18 ~ "Under 18",
    AGE >= 18 & AGE <= 25 ~ "18-25",
    AGE >= 26 & AGE <= 35 ~ "26-35",
    AGE >= 36 & AGE <= 50 ~ "36-50",
    AGE > 50 ~ "Above 50",
    TRUE ~ "Unknown"
  )) %>%
  filter(Age_Group != "Unknown" & SEAT_POS == "Rider") %>%
  mutate(Age_Group = factor(Age_Group, levels = c("Under 18", "18-25", "26-35", "36-50", "Above 50")))
```

### Drinking

```{r}
moto_drink_total <- moto_data %>%
  filter(., DRINKING_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

moto_drink_death <- moto_data %>%
  filter(INJURY == "Death" & DRINKING_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

# Create a bar chart
ggplot() +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Total"), 
    data = moto_drink_total,
    stat = "identity", show.legend = FALSE
  ) +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Death"), 
    data = moto_drink_death,
    stat = "identity", show.legend = FALSE
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 10)  # Generate 10 evenly spaced breaks
  ) +
  scale_fill_manual(
    values = c("Total" = "#1fbcff", "Death" = "#ff661f"),  # Assign colors to categories
    labels = c("Total Count", "Deaths")              # Legend labels
  ) +
  labs(
    title = "Accidents Across Age Groups",
    subtitle = "Drinking and Death",
    x = "Age Group",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
  )

ggsave("moto_drink_death.png", path = "plots", dpi = 350)
```
### Drugs

```{r}
moto_drugs_total <- moto_data %>%
  filter(., DRUGS_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

moto_drugs_death <- moto_data %>%
  filter(INJURY == "Death" & DRUGS_FLAG == "Yes") %>%
  group_by(Age_Group) %>%
  summarise(Count = n(), .groups = "drop")

# Create a bar chart
ggplot() +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Total"), 
    data = moto_drugs_total,
    stat = "identity", show.legend = FALSE
  ) +
  geom_bar(
    mapping = aes(x = Age_Group, y = Count, fill = "Death"), 
    data = moto_drugs_death,
    stat = "identity", show.legend = FALSE
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 10)  # Generate 10 evenly spaced breaks
  ) +
  scale_fill_manual(
    values = c("Total" = "#1fbcff", "Death" = "#ff661f"),  # Assign colors to categories
    labels = c("Total Count", "Deaths")              # Legend labels
  ) +
  labs(
    title = "Accidents Across Age Groups",
    subtitle = "Drugs and Death",
    x = "Age Group",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
  )

ggsave("moto_drugs_death.png", path = "plots", dpi = 350)
```




# Pie Charts

```{r}
View(vehicles %>% select(HARM_EV_TYPE) %>% distinct())
```

```{r}

harm_event_fil <- c("Pedestrian", "Pedalcyclist", "Fell/Jumped from Vehicle", "Unknown", "Non-Motorist on Personal Conveyance")

# Summarize the data to count occurrences of HARM_EV_TYPE for each INJURY
harm_event_summary <- vehicles %>%
  filter(., INJURY != "Dead Prior Crash" & !HARM_EV_TYPE %in% harm_event_fil) %>%
  group_by(INJURY, HARM_EV_TYPE) %>%
  summarise(count = n(), .groups = "drop")

# Filter to get the top 5 HARM_EV_TYPE for each INJURY category
top_5_harm_events <- harm_event_summary %>%
  group_by(INJURY) %>%
  slice_max(order_by = count, n = 5) %>%
  ungroup()

# Visualize the data using a faceted bar chart
ggplot(top_5_harm_events, aes(x = reorder(HARM_EV_TYPE, -count), y = count, fill = INJURY)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    title = "Top 5 Harm Event Types by Injury Category",
    x = "Harm Event Type",
    y = "Count"
  ) +
  facet_wrap(~INJURY, scales = "free") +  # Facet by INJURY
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    strip.text = element_text(size = 10)  # Adjust facet label size
  )

```
